---
title: "BIS15L_AI_hmwk"
author: "Your Name"
date: "`r Sys.Date()`"
output: 
  html_document: 
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
This homework is a bit less traditional this week. There are some errors in the code below. In addition add notes and titles to the code chunks that explains what the code chunck does. Use AI tools, notes from class, or any other resources to help fix the errors. Feel free to edit any plots to your preference. Don't forget to answer the two questions at the bottom. Upon completion, you should not have to '#' out any of the code chunks.

Push your final work to your repository. Your final lab report should be organized, clean, and run free from errors. Remember, you must remove the `#` for the included code chunks to run. Be sure to add your name to the author header above.   

Again make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  


```{r}
library(tidyverse)
library(janitor)
library(dplyr)
```


1.
```{r}
blood_count <- read_csv("data/blood_count_dataset.csv")%>% clean_names()
```

2. 
```{r}
glimpse(blood_count)

```

3. 
```{r}
blood_count %>%
  group_by(gender) %>%
  summarize(min = min(red_blood_cells),
            mean = mean(red_blood_cells), # Fixed typo: changed "maen" to "mean"
            max = max(red_blood_cells))

```


4. 
```{r}
options(scipen = 999, digits = 2)

blood_count %>%
  filter(gender == "male") %>% # Use == for comparison
  select(3:9) %>% # Use parentheses for select()
  colMeans(na.rm = TRUE)

```

5. 
```{r}
blood_count %>% 
  filter(gender == "Female") %>% # Use == for comparison and correct pipeline operator
  select(3:9) %>% # Use parentheses ()
  colMeans(na.rm = TRUE) # Use TRUE

```


6. 
```{r}


blood_count %>%
  pivot_longer(cols = hemoglobin:mchc, names_to = "Test", values_to = "Values") %>%
  ggplot(aes(x = gender, y = Values, fill = gender)) +
  geom_boxplot() +
  facet_wrap(~ Test, scales = "free") + 
  labs(title = "Boxplot of Blood Count Tests",
       x = "Gender",
       y = "Values")



```

7. 
```{r}
library(dplyr)

grouped_bc <- blood_count %>%
  mutate(age_Group = case_when(
    age >= 20 & age <= 29 ~ "20s",
    age >= 30 & age <= 39 ~ "30s",
    age >= 40 & age <= 49 ~ "40s",
    age >= 50 & age <= 59 ~ "50s",
    age >= 60 & age <= 70 ~ "60s",
    TRUE ~ "Other")) # Handling ages outside the specified ranges
# Display the first few rows of the modified dataframe to verify the changes
head(grouped_bc)


```

8. 
```{r}
grouped_bc %>%
  pivot_longer(cols = 3:9, names_to = "Test", values_to = "Values") %>%
  ggplot(aes(x = age_Group, y = Values, fill = age_Group)) +
  geom_boxplot() +
  facet_wrap(~ Test, scales = "free") +  # Facet by the "Test" variable
  labs(title = "Boxplot of Blood Count Tests by Age Group",
       x = "Age Group",
       y = "Values")

```
9. What are two drawbacks and two benefits of AI?
AI enhances efficiency and decision-making by automating tasks and analyzing data more accurately than humans, which can lead to significant advancements across various fields. However, it poses risks such as job displacement due to automation and ethical concerns, including privacy issues and potential biases in decision-making processes. Balancing AI's benefits with its drawbacks requires careful management and regulation to ensure equitable and responsible use.


10. Do you think you will use AI for coding in the future? Why or why not?

Its very helpful with catching errors but I feel like a lot of the times above when I asked it find errors it took longer to find them than I did. Other times though, it was very efficient. Therefore, if I was having trouble with finding errors and solving my code, I would definitley use it as a resource. 

## Push your final code to GitHub!
Please be sure that you check the `keep md` file in the knit preferences. 